Preschoolers form conventional pacts with each other to communicate about novel shapes

### 2025-01-21 write-ups/kid_tangrams_cogsci_2025/kid_tangrams_cogsci_2025.Rmd
refactor ms to pull out prep and model code into helpers

# experiment 1 imports
transcript_1 <- read_csv(here(expt_1, "timed_transcript.csv"))
dat_1 <- read_csv(here(expt_1, "clean_data.csv")) |> rename(trial = trialNum)
exclusion_1 <- transcript_1 %>%
  filter(!is.na(description)) %>%
  filter(role == "S") %>%
  select(game, trial) %>%
  unique() %>%
  mutate("S_talked" = 1)

words_1 <- transcript_1 |>
  filter(!is.na(description)) |>
  mutate(words = description |> str_count("\\S+")) |>
  filter(role == "S") |>
  group_by(game, trial) |>
  summarize(words = sum(words))

clean_1 <- dat_1 %>%
  left_join(exclusion_1) %>%
  filter(S_talked == 1) |>
  left_join(words_1) |>
  mutate(type = case_when(
    trial < 2 ~ "practice",
    trial < 6 ~ "block 1",
    trial < 10 ~ "block 2",
    trial < 14 ~ "block 3"
  ))

sims_1 <- read_rds(here(expt_1, "similarities.rds")) |>
  as_tibble() |>
  inner_join(clean_1 |> select(game1 = game, trial1 = trial, target1 = target, speaker1 = speaker)) |>
  inner_join(clean_1 |> select(game2 = game, trial2 = trial, target2 = target, speaker2 = speaker)) |>
  filter(target1 == target2) |>
  filter(trial1 > 1) |>
  filter(trial2 > 1) |>
  mutate(
    trial1 = trial1 + 2,
    trial2 = trial2 + 2
  ) |>
  mutate(block1 = trial1 %/% 4, block2 = trial2 %/% 4) |>
  mutate(
    same_speaker = (speaker1 == speaker2) |> as.numeric(),
    same_game = (game1 == game2) |> as.numeric(),
    later = ifelse(block1 > block2, block1, block2),
    earlier = ifelse(block1 > block2, block2, block1)
  ) |>
  mutate(target = target1)

# expt 2 imports
transcript_2 <- read_csv(here(expt_2, "transcripts.csv"))
link <- read_csv(here(expt_2, "link_transcripts.csv"))

exclude_2 <- transcript_2 |>
  filter(!is.na(echo)) |>
  select(gameId, trial)

data_2 <- read_csv(here(expt_2, "clean_data.csv")) |>
  rename(trial = trialNum) |>
  anti_join(exclude_2)

transcript_2 <- transcript_2 |>
  anti_join(exclude_2) |>
  filter(!is.na(description))

words_2 <- transcript_2 |>
  mutate(words = description |> str_count("\\S+")) |>
  filter(role == "S") |>
  group_by(gameId, gameConfig, trial) |>
  summarize(words = sum(words))

clean_2 <- data_2 |>
  left_join(words_2) |>
  filter(!is.na(words)) |>
  mutate(type = case_when(
    trial < 4 ~ "practice",
    trial < 8 ~ "block 1",
    trial < 12 ~ "block 2",
    trial < 16 ~ "block 3",
    trial < 20 ~ "block 4"
  )) |>
  filter(!is.na(response)) |>
  mutate(correct = as.numeric(correct))

sims_2 <- read_rds(here(expt_2, "similarities.rds")) |>
  as_tibble() |>
  inner_join(clean_2 |> select(game1 = gameId, trial1 = trial, target1 = target, speaker1 = speaker)) |>
  inner_join(clean_2 |> select(game2 = gameId, trial2 = trial, target2 = target, speaker2 = speaker)) |>
  filter(target1 == target2) |>
  mutate(block1 = trial1 %/% 4, block2 = trial2 %/% 4) |>
  mutate(
    same_speaker = (speaker1 == speaker2) |> as.numeric(),
    same_game = (game1 == game2) |> as.numeric(),
    later = ifelse(block1 > block2, block1, block2),
    earlier = ifelse(block1 > block2, block2, block1)
  ) |>
  mutate(target = target1)
```

```{r, eval=F}
library(tidybayes)

save_summary <- function(model) {
  intervals <- gather_draws(model, `b_.*`, regex = T) %>% mean_qi()

  stats <- gather_draws(model, `b_.*`, regex = T) %>%
    mutate(above_0 = ifelse(.value > 0, 1, 0)) %>%
    group_by(.variable) %>%
    summarize(pct_above_0 = mean(above_0)) %>%
    left_join(intervals, by = ".variable") %>%
    mutate(
      lower = .lower,
      upper = .upper,
      Term = str_sub(.variable, 3, -1),
      Estimate = .value
    ) %>%
    select(Term, Estimate, lower, upper)

  stats
}

do_model <- function(path) {
  model <- read_rds(here(mod_loc, path))
  save_summary(model) |> write_rds(here(mod_loc, "summary", path))
  model$formula |> write_rds(here(mod_loc, "formulae", path))
  print(summary(model))
}

mods <- list.files(path = here(mod_loc), pattern = ".*rds") |> walk(~ do_model(.))
```

```{r}
stats <- function(model, row, decimal = 2) {
  model <- model |>
    mutate(
      Estimate = round(Estimate, digits = decimal),
      Lower = round(lower, digits = decimal),
      Upper = round(upper, digits = decimal),
      `Credible Interval` = str_c("[", Lower, ", ", Upper, "]")
    ) |>
    select(Term, Estimate, `Credible Interval`)
  str_c(model[row, 1], ": ", model[row, 2], " ", model[row, 3])
}

stats_text <- function(model, row, decimal = 2) {
  model <- model |>
    mutate(
      Estimate = round(Estimate, digits = decimal) |> formatC(format='f', digits=decimal ),
      Lower = round(lower, digits = decimal) |> formatC(format='f', digits=decimal ),
      Upper = round(upper, digits = decimal) |> formatC(format='f', digits=decimal ),
      `Credible Interval` = str_c("[", Lower, ", ", Upper, "]")
    ) |>
    select(Term, Estimate, `Credible Interval`)
  str_c(model[row, 2], "  ", model[row, 3])
}

form <- function(model_form) {
  dep <- as.character(model_form$formula[2])
  ind <- as.character(model_form$formula[3])

  str_c(dep, " ~ ", ind) |>
    str_replace_all(" ", "") |>
    str_replace_all("\\*", " $\\\\times$ ") |>
    str_replace_all("\\+", "&nbsp;+ ") |>
    str_replace_all("~", "$\\\\sim$ ")
}
```

@@

<!-- intro -->

@@

One test of referential communication skill is *iterated reference games* where one describes abstract shapes from an array of images repeatedly to the same partner [@clark1986; @krauss1964; @hawkins2020b; @boyce2024]. Over repetition, features of the initial verbose descriptions are conventionalized as each pair comes to agree on a shared understanding of how to label each image. Success at this task requires mastery of a number of linguistic and communicative skills, including producing adequate initial descriptions, monitoring for comprehension, asking for clarification, and appropriately using the shared conversation history to inform referring expressions in later rounds. 
<!--Indeed, even young children must make use of these skills in order to achieve success in the task. However, there is debate on the developmental trajectories of these communicative skills. --> Studying how children play iterated reference games can provide insight into the developmental trajectory of the ability to produce referential expressions in order to achieve joint understanding.

A classic study suggests that 4-5-year old preschoolers struggle with child-child referential communication [@glucksberg1966]. In their paradigm, one child was given a set of 6 blocks in a specific order. Their task was to describe the images on the block so their partner could pick out the corresponding block. As children described and selected blocks, they stacked them on pegs. While 4-5-year old children were successful on practice trials with familiar shapes and visual access to each other's blocks, children failed on critical trials where the blocks had drawings of abstract shapes and there was no visual access. Even after multiple rounds with the same images, children were not able to correctly order the blocks. <!--@glucksberg1966 attributed young children's failure to the children’s use of idiosyncratic referring expressions unique to their own experiences, making it difficult for the pair to converge on shared descriptions.-->
Similar experiments with older children indicated a gradual improvement through adolescence both for initial accuracy and for the increase in accuracy across repetitions. Still, even the 9th grade sample was noticeably worse than the sample of adult college students [@glucksberg1967; @krauss1969]. That the task was difficult even among teenagers suggests that the complexity of the stacking paradigm and the large number of potential targets present significant task demands that might prevent accurate measurement of children's abilities.

@@

<!--(Note: we should talk about whether it makes more sense to do all the methods and then all the results versus the current expt 1 than expt 2 layout.) -->

Our goal for experiment 1 was to test young children's ability to coordinate on descriptions to abstract shapes that their partner could understand. Young children can be very sensitive to task demands and cognitive load (CITATIONS), so we adapted the experimental framework from @leung2024, and further simplified it by reducing the total pool of targets and the number of trials. This experiment was pre-registered at https://osf.io/kcv8j.

### Participants
4 and 5-year old children were recruited from a university preschool during the school day. Children played with another child from the same class. Experiment 1 was conducted between June and August 2023. Pairs of children were included in analyses if they completed at least 8 of the 12 critical trials. We had 19 complete games and 1 included incomplete game. Of the 40 children, 21 were girls, and the median age was 57 months, with a range of 48-70 months.

```{r, eval=F}
# demo <- read_csv(here("demographics.csv")) |>
#   filter(included == "x") |>
#   mutate(date_test = mdy(date_test), date_birth = mdy(date_birth)) |>
#   select(date_test, date_birth, gender) |>
#   mutate(
#     age = date_test - date_birth,
#     age = as.numeric(age) / 365.25 * 12
#   ) |> 
#   mutate(expt=ifelse(date_test>ymd("2024-01-01"), "expt2", "expt1")) |> 
#   group_by(expt) |> 
#   summarize(female=sum(gender=="female"),
#             total=n(),
#             median=median(age),
#             min=min(age),
#             max=max(age))
```

@@

The experimenters running the game did not volunteer descriptions, but they did scaffold the interaction, prompting children to describe the images, and sometimes repeating children's statements. The entire interaction was video-recorded.

@@

Children's selections and the time to selection were recorded from the experiment software. Children's descriptions were transcribed from the video-recording, using Whisper [@radford2022] for the first pass and then hand-corrected by experimenters. Transcripts were hand-annotated for when each trial started, who said each line, and what referential descriptions were used.

@@

```{r fig-accuracy, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Children's accuracy at selecting the correct target over time. Experiment 1 had 3 blocks (12 total critical trials) and experiment 2 had 4 blocks (16 critical trials). Error bars are bootstrapped 95% CIs with a linear trend line overlaid. \\label{acc}" }
clean_2 |>
  filter(type != "practice") |>
  mutate(trial = trial - 3, expt = "Expt 2") |>
  bind_rows(clean_1 |> filter(type != "practice") |> mutate(trial = trial - 1, expt = "Expt 1")) |>
  ggplot(aes(x = trial, y = correct)) +
    geom_smooth(method = "lm", color = "black") +
  stat_summary(aes(color = type), fun.data = "mean_cl_boot") +
  #stat_summary(fun.data = "mean_cl_boot", geom = "linerange", color = "black") +
  scale_color_manual(values = time_palette) +
  # geom_bar(position = position_dodge(.9), color = "black", fill = "#19ADDE", width = .7) +
  # geom_text(aes(label = percent), hjust = 1.6, color = "white", size = 3) +
  xlab("Trial") +
  ylab("Accuracy") +
  geom_vline(xintercept = 4.5, lty="dashed")+
    geom_vline(xintercept = 8.5, lty="dashed")+
    geom_vline(xintercept = 12.5, lty="dashed")+
  scale_x_continuous(breaks = seq(1, 16, 4), expand = c(0.02, 0.02)) +
  scale_y_continuous(breaks = seq(0, 1, .2), expand = c(0, 0.04)) +
  coord_cartesian(ylim = c(0.3,1))+
  geom_hline(yintercept = .5) +
  facet_grid(expt ~ .) +
  theme(
    legend.position = "none",
    axis.ticks.x=element_blank(),
    axis.text.x=element_blank(),
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

```{r, eval=F}
acc_priors <- c(
  set_prior("normal(0,1)", class = "b"),
  set_prior("normal(0,1)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

acc_mod_data_1 <- clean_1 |>
  mutate(
    correct.num = ifelse(correct == TRUE, 1, 0),
    trial.num = trial - 2
  ) |>
  filter(type != "practice")
acc_mod_1 <- brm(correct.num ~ trial.num + (trial.num | game) + (1 | target),
  family = bernoulli(link = "logit"),
  data = acc_mod_data_1,
  file = here(mod_loc, "acc_1.rds"),
  prior = acc_priors,
  control = list(adapt_delta = .95)
)
```

@@

Our primary measure of interest was whether children could accurately communicate the intended target. <!-- as prior work is often interpreted as indicating the children at kindergarten age cannot communicate about abstract shapes successfully [@glucksberg1966]. -->  As shown in Figure \ref{acc}, children were above chance in their selections. To test for changes in accuracy over time, we fit a Bayesian mixed effects logistic regression predicting accuracy (`r acc_form |> form()`). Children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_1,1)`) and their accuracy slightly increased over the game (OR of one trial later: `r stats_text(acc_mod_1, 2)`).

```{r, eval=F}
speed_priors <- c(
  set_prior("normal(60,100)", class = "Intercept"),
  set_prior("normal(0,20)", class = "b"),
  set_prior("normal(0,20)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

speed_mod_data_1 <- clean_1 |>
  mutate(
    time.sec = time / 1000,
    trial.num = trial - 2
  ) |>
  filter(type != "practice")

speed_mod_1 <- brm(time.sec ~ trial.num + (trial.num | game) + (1 | target),
  data = speed_mod_data_1,
  file = here(mod_loc, "speed_1.rds"),
  prior = speed_priors,
  control = list(adapt_delta = .95)
)
```

@@

<!--[less important, could drop] --> As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects linear regression predicting the time to selection in seconds: `r speed_form |> form()`. The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)` seconds / trial).

@@

```{r, eval=F}
description_priors <- c(
  set_prior("normal(5,10)", class = "Intercept"),
  set_prior("normal(0,5)", class = "b"),
  set_prior("normal(0,5)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

description_mod_data_1 <- clean_1 |>
  mutate(
    trial.num = trial - 2
  ) |>
  filter(type != "practice")

description_mod_1 <- brm(words ~ trial.num + (trial.num | game) + (1 | target),
  data = description_mod_data_1,
  file = here(mod_loc, "description_1.rds"),
  prior = speed_priors,
  control = list(adapt_delta = .95)
)
```

```{r description, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Length of referential expression (in words) produced by the speaker each trial. Grey dots are individual data points, colored dots are per trial means with bootstrapped 95% CIs. \\label{length}" }
clean_2 |>
  filter(type != "practice") |>
  mutate(trial = trial - 3, expt = "Expt 2") |>
  bind_rows(clean_1 |> filter(type != "practice") |> mutate(trial = trial - 1, expt = "Expt 1")) |>
  ggplot(aes(x = trial, y = words, color = type)) +
  geom_jitter(alpha = .5, color = "grey") +
  stat_summary(fun.data = "mean_cl_boot") +
  scale_x_continuous(breaks = seq(1, 20, 1), expand = c(0, 0)) +
  scale_color_manual(values = time_palette) +
  geom_smooth(method = "lm", color = "black") +
  xlab("Trial") +
  ylab("Words") +
  facet_grid(expt ~ .) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

In iterated reference games with adults, description lengths usually shorten over repeated references [@clark1986; @hawkins2020b; @boyce2024]. We were curious if children's descriptions would display the same trend, so we ran a Bayesian mixed effects linear regression predicting the number of words in the description the "teller" produced: `r description_form |> form()`. On the first critical trial, descriptions averaged `r stats_text(description_mod_1, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_1, 2)` words per trial, Figure \ref{length}).

@@

```{r, eval=F}
conv_priors <- c(
  set_prior("normal(.5,.2)", class = "Intercept"),
  set_prior("normal(0,.1)", class = "b"),
  set_prior("normal(0,.05)", class = "sd")
)

sim_across_between_data_1 <- sims_1 |>
  mutate(same_game = case_when(
    game1 == game2 ~ 1,
    T ~ 0
  )) |>
  mutate(same_speaker = ifelse(speaker1 == speaker2, 1, 0))

sim_across_between_mod_1 <- brm(sim ~ same_game + same_speaker + (1 | target),
  data = sim_across_between_data_1,
  file = here(mod_loc, "sim_across_between_1.rds"),
  prior = conv_priors,
  control = list(adapt_delta = .95)
)
```

```{r sims, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Semantic similarity between referential descriptions from the speaker to the same target under different circumstances. Different games refers to similarities across two speakers from different games, same game to similarities across the two participants in one game, and same speaker to descriptions from the same person in different blocks. Dots are means and lines are bootstrapped 95% CIs.  \\label{sim}" }
sims_2 |>
  mutate(source = case_when(
    speaker1 == speaker2 ~ "same speaker",
    game1 == game2 ~ "same game",
    T ~ "different games"
  )) |>
  mutate(expt = "Expt 2") |>
  bind_rows(sims_1 |>
    mutate(source = case_when(
      speaker1 == speaker2 ~ "same speaker",
      game1 == game2 ~ "same game",
      T ~ "different games"
    )) |> mutate(expt = "Expt 1")) |>
  ggplot(aes(x = source, y = sim, color = target1)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  labs(y = "Cosine similarity", x = "Comparing two utterances from ...", color = "target image") +
  facet_grid(expt ~ .) +
  scale_color_solarized() +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

```{r, include=F}
sims_1 |>
  filter(game1 == game2) |>
  mutate(later = ifelse(block1 > block2, block1, block2), earlier = ifelse(block1 > block2, block2, block1)) |>
  filter(later == 3) |>
  mutate(same_speaker = ifelse(speaker1 == speaker2, "same speaker", "different speaker")) |>
  ggplot(aes(as.character(earlier), sim, color = target1)) +
  geom_point(alpha = .1, position = position_dodge(width = .2)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  facet_wrap(~same_speaker) +
  labs(x = "Earlier block to last block", y = "cosine similarity", color = "target") +
  theme(legend.position = "bottom")
```

```{r include=F}
sims_1 |>
  filter(game1 == game2) |>
  mutate(later = ifelse(block1 > block2, block1, block2), earlier = ifelse(block1 > block2, block2, block1)) |>
  filter(later == earlier + 1) |>
  mutate(same_speaker = ifelse(speaker1 == speaker2, "same speaker", "different speaker")) |>
  ggplot(aes(as.character(earlier), sim, color = target1)) +
  geom_point(alpha = .1, position = position_dodge(width = .2)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  facet_wrap(~same_speaker) +
  labs(x = "Earlier block to *next* block", y = "cosine similarity", color = "target") +
  theme(legend.position = "bottom")
```

```{r, include=F}
sims_1 |>
  filter(game1 != game2) |>
  mutate(later = ifelse(block1 > block2, block1, block2), earlier = ifelse(block1 > block2, block2, block1)) |>
  filter(later == earlier) |>
  ggplot(aes(as.character(earlier), sim, color = target1)) +
  geom_point(alpha = .01, position = position_dodge(width = .2)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  labs(x = "Cross game similarity in block ...", y = "cosine similarity", color = "target") +
  theme(legend.position = "bottom")
```

While description length is used as a proxy for measuring convention formation, better measurements for convergence look at the actual content of the utterances, which does not always track description length [@boyce2024]. Following @boyce2024, we use Sentence-BERT to embed the descriptions in a semantic vector space, and then use the cosine between embeddings as a measure of semantic similarity between descriptions [@reimers2019].

We compared the semantic similarities between descriptions of the same target based on who produced the description. Our question was whether the two children within the same game produced descriptions that are more similar to each other than two children in different games.

Different children in the same game produced more similar descriptions than children in different games, although the descriptions were less similar than descriptions from the same child in different rounds (Figure \ref{sim}). We modeled this as a Bayesian mixed effects linear regression predicting similarity `r sim_across_between_form |> form()`. Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_1_results |> stats_text(3,3)`). The big differences in descriptions between games compared to within games is a measure of partner sensitivity.

@@

### Participants
Experiment 2 was run between March and August of 2024, at the same university preschool as experiment 1. No children participated in both experiments. 30 pairs of children completed all 16 critical trials, and 1 pair of children completed between 8 and 16 critical trials. Our target age range was 4 and 5 year olds, but one almost 4-year-old was unintentionally included. Of the 62 children, 30 were girls, and the children had a median age of 56 months, and a range of 45-69 months.

### Materials
The same 4 critical images were used as in experiment 1, although this time, children saw these images 4 times. In response to some children struggling with the abrupt switch from familiar to non-nameable shapes, we introduced more practice trials for experiment 2. We used a total of 4 practice trials to provide a gradient from familiar shapes to less recognizable, blockier shapes (Figure \ref{game}D).

@@

### Data processing
Data was processed in the same way as experiment 1. After excluding trials where children did not give a description or where the experimenter echoed a child's description, we had `r clean_2 |> filter(!is.na(repNum)) |> nrow()` trials total.

@@

We report the same set of analyses and model results as in Experiment 1 as well as additional analyses of how the semantic content of children's descriptions changes over time. Children used a wide range of descriptions successfully, and some example descriptions from Experiment 2 are shown in Table \ref{example}.

@@

```{r, eval=F}
acc_priors <- c(
  set_prior("normal(0,1)", class = "b"),
  set_prior("normal(0,1)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

acc_mod_data <- clean_2 |>
  mutate(
    correct.num = ifelse(correct == TRUE, 1, 0),
    trial.num = trial - 4
  ) |>
  filter(type != "practice")
acc_mod <- brm(correct.num ~ trial.num + (trial.num | gameId) + (1 | target),
  family = bernoulli(link = "logit"),
  data = acc_mod_data,
  file = here(mod_loc, "acc_2.rds"),
  prior = acc_priors,
  control = list(adapt_delta = .95)
)
```

@@

```{r}
speed_priors <- c(
  set_prior("normal(60,100)", class = "Intercept"),
  set_prior("normal(0,20)", class = "b"),
  set_prior("normal(0,20)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

speed_mod_data_2 <- clean_2 |>
  mutate(
    time.sec = time / 1000,
    trial.num = trial - 2
  ) |>
  filter(type != "practice")

speed_mod_2 <- brm(time.sec ~ trial.num + (trial.num | gameId) + (1 | target),
  data = speed_mod_data_2,
  file = here(mod_loc, "speed_2.rds"),
  prior = speed_priors,
  control = list(adapt_delta = .95)
)
```

@@

```{r, eval=F}
description_priors <- c(
  set_prior("normal(5,10)", class = "Intercept"),
  set_prior("normal(0,5)", class = "b"),
  set_prior("normal(0,5)", class = "sd"),
  set_prior("lkj(1)", class = "cor")
)

description_mod_data_2 <- clean_2 |>
  mutate(
    trial.num = trial - 2
  ) |>
  filter(type != "practice")

description_mod_2 <- brm(words ~ trial.num + (trial.num | gameId) + (1 | target),
  data = description_mod_data_2,
  file = here(mod_loc, "description_2.rds"),
  prior = speed_priors,
  control = list(adapt_delta = .95)
)
```

@@

## Pooling across experiments

>>>>

source(here("code/paper_helper.R"))
```

@@

@@

One test of referential communication skill is *iterated reference games* where one describes abstract shapes from an array of images repeatedly to the same partner [@clark1986; @krauss1964; @hawkins2020b; @boyce2024]. Over repetition, features of the initial verbose descriptions are conventionalized as each pair comes to agree on a shared understanding of how to label each image. Success at this task requires mastery of a number of linguistic and communicative skills, including producing adequate initial descriptions, monitoring for comprehension, asking for clarification, and appropriately using the shared conversation history to inform referring expressions in later rounds. 
Studying how children play iterated reference games can provide insight into the developmental trajectory of the ability to produce referential expressions in order to achieve joint understanding.

A classic study suggests that 4-5-year old preschoolers struggle with child-child referential communication [@glucksberg1966]. In their paradigm, one child was given a set of 6 blocks in a specific order. Their task was to describe the images on the block so their partner could pick out the corresponding block. As children described and selected blocks, they stacked them on pegs. While 4-5-year old children were successful on practice trials with familiar shapes and visual access to each other's blocks, children failed on critical trials where the blocks had drawings of abstract shapes and there was no visual access. Even after multiple rounds with the same images, children were not able to correctly order the blocks. 
Similar experiments with older children indicated a gradual improvement through adolescence both for initial accuracy and for the increase in accuracy across repetitions. Still, even the 9th grade sample was noticeably worse than the sample of adult college students [@glucksberg1967; @krauss1969]. That the task was difficult even among teenagers suggests that the complexity of the stacking paradigm and the large number of potential targets present significant task demands that might prevent accurate measurement of children's abilities.

@@

Our goal for experiment 1 was to test young children's ability to coordinate on descriptions to abstract shapes that their partner could understand. Young children can be very sensitive to task demands and cognitive load [@turan-kucuk2024; @keen2003; @carruthers2013], so we adapted the experimental framework from @leung2024, and further simplified it by reducing the total pool of targets and the number of trials. This experiment was pre-registered at https://osf.io/kcv8j.

### Participants

4 and 5-year old children were recruited from a university preschool during the school day. Children played with another child from the same class. Experiment 1 was conducted between June and August 2023. Pairs of children were included in analyses if they completed at least 8 of the 12 critical trials. We had 19 complete games and 1 included incomplete game. Of the 40 children, 21 were girls, and the median age was 57 months, with a range of 48-70 months.

@@

The experimenters running the game did not volunteer descriptions, but they did scaffold the interaction, prompting children to describe the images, and sometimes repeating children's statements (especially when utterances were inaudible or the child did not respond immediately; this aspect of the procedure was modified in Experiment 2). The entire interaction was video-recorded.

@@

Children's selections and the time to selection were recorded from the experiment software. Children's descriptions were transcribed from the video using Whisper [@radford2022] for the first pass and then hand-corrected by experimenters. Transcripts were hand-annotated for when each trial started, who said each line, and what referential descriptions were used.

@@

### Accuracy and Speed

```{r fig-accuracy, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Children's accuracy at selecting the correct target over time. Experiment 1 had 3 blocks (12 total critical trials) and experiment 2 had 4 blocks (16 critical trials). Error bars are bootstrapped 95% CIs with a linear trend line overlaid. \\label{acc}" }
clean_2 |>
  filter(type != "practice") |>
  mutate(trial = trial - 3, expt = "Expt 2") |>
  bind_rows(clean_1 |> filter(type != "practice") |> mutate(trial = trial - 1, expt = "Expt 1")) |>
  ggplot(aes(x = trial, y = correct)) +
    geom_smooth(method = "lm", color = "black") +
  stat_summary(aes(color = type), fun.data = "mean_cl_boot") +
  #stat_summary(fun.data = "mean_cl_boot", geom = "linerange", color = "black") +
  scale_color_manual(values = time_palette) +
  # geom_bar(position = position_dodge(.9), color = "black", fill = "#19ADDE", width = .7) +
  # geom_text(aes(label = percent), hjust = 1.6, color = "white", size = 3) +
  xlab("Trial") +
  ylab("Accuracy") +
  geom_vline(xintercept = 4.5, lty="dashed")+
    geom_vline(xintercept = 8.5, lty="dashed")+
    geom_vline(xintercept = 12.5, lty="dashed")+
  scale_x_continuous(breaks = seq(1, 16, 1), expand = c(0.02, 0.02)) +
  scale_y_continuous(breaks = seq(0, 1, .2), expand = c(0, 0.04)) +
  coord_cartesian(ylim = c(0.3,1))+
  geom_hline(yintercept = .5) +
  facet_grid(expt ~ .) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

Our primary measure of interest was whether children could accurately communicate the intended target. <!-- as prior work is often interpreted as indicating the children at kindergarten age cannot communicate about abstract shapes successfully [@glucksberg1966]. -->  As shown in Figure \ref{acc}, children generally accurate in their selections. To test for changes in accuracy over time, we fit a Bayesian mixed effects logistic regression predicting accuracy.^[`r acc_form |> form()`] Children's accuracy was above chance (Odds Ratio: `r stats_text(acc_mod_1,1)`), and their accuracy slightly increased over the game (OR of one trial later: `r stats_text(acc_mod_1, 2)`). This level of accuracy is generally in-line with accuracies from 4-year-olds playing with their parents in @leung2024 and indicates that children can understand and succeed at the task.

@@

As another measure of children's performance, we looked at how long each trial took to see if children were getting faster over time. We ran a Bayesian mixed effects linear regression predicting the time to selection in seconds.^[`r speed_form |> form()`] The first critical trial averaged `r stats_text(speed_mod_1, 1)` seconds, and children got faster over time (`r stats_text(speed_mod_1, 2)` seconds / trial). This is a measure of efficiency as children were able to achieve the same accuracy levels in less time.

@@

```{r description, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Length of referential expression (in words) produced by the teller each trial. Grey dots are individual data points, colored dots are per trial means with bootstrapped 95% CIs. \\label{length}" }
clean_2 |>
  filter(type != "practice") |>
  mutate(trial = trial - 3, expt = "Expt 2") |>
  bind_rows(clean_1 |> filter(type != "practice") |> mutate(trial = trial - 1, expt = "Expt 1")) |>
  ggplot(aes(x = trial, y = words, color = type)) +
  geom_jitter(alpha = .5, color = "grey") +
  stat_summary(fun.data = "mean_cl_boot") +
  scale_x_continuous(breaks = seq(1, 20, 1), expand = c(0, 0)) +
    geom_vline(xintercept = 4.5, lty="dashed")+
    geom_vline(xintercept = 8.5, lty="dashed")+
    geom_vline(xintercept = 12.5, lty="dashed")+
  scale_color_manual(values = time_palette) +
  geom_smooth(method = "lm", color = "black") +
  xlab("Trial") +
  ylab("Words") +
  facet_grid(expt ~ .) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

In iterated reference games with adults, description lengths usually shorten over repeated references [@clark1986; @hawkins2020b; @boyce2024]. We were curious if children's descriptions would display the same trend, so we ran a Bayesian mixed effects linear regression predicting the number of words in the description the "teller" produced.^[`r description_form |> form()`] On the first critical trial, descriptions averaged `r stats_text(description_mod_1, 1)` words and description length was relatively stable over time (change of `r stats_text(description_mod_1, 2)` words per trial, Figure \ref{length}). Thus, while children were taking less time per trial, their speed-up was not from shorter utterances, instead they were understanding the game better and taking less time to initiate their descriptions and make their selections.

Some examples to illustrate the variety of effective descriptions children employed are shown in Table \ref{example}; these specific examples are from Experiment 2, but are also representative of the types of descriptions produced in Experiment 1.

@@

```{r sims, fig.env="figure", fig.pos = "t", fig.align = "center", out.width="100%", fig.width=5, fig.height=3, fig.cap = "Semantic similarity between referential descriptions from the teller to the same target under different circumstances. Different games refers to similarities across two children from different games, same game to similarities across the two participants in one game, and same teller to descriptions from the same person in different blocks. Dots are means and lines are bootstrapped 95% CIs.  \\label{sim}" }
sims_2 |>
  mutate(source = case_when(
    speaker1 == speaker2 ~ "same teller",
    game1 == game2 ~ "same game",
    T ~ "different games"
  )) |>
  mutate(expt = "Expt 2") |>
  bind_rows(sims_1 |>
    mutate(source = case_when(
      speaker1 == speaker2 ~ "same teller",
      game1 == game2 ~ "same game",
      T ~ "different games"
    )) |> mutate(expt = "Expt 1")) |>
  ggplot(aes(x = source, y = sim, color = target1)) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(width = .2)) +
  labs(y = "Cosine similarity", x = "Comparing two utterances from ...", color = "Target") +
  facet_grid(expt ~ .) +
  scale_color_solarized() +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.background = element_blank()
  )
```

@@

While description length is used as a proxy for measuring convention formation, semantic overlap between utterances does not always track description length. @boyce2024 introduced a more sensitive measure of semantic convergence.  By comparing the content of utterances using word embeddings, @boyce2024 traced when how similarities between utterances in different rounds and games

Following @boyce2024, we use Sentence-BERT to embed the descriptions in a semantic vector space, and then use the cosine between embeddings as a measure of semantic similarity between descriptions [@reimers2019].

We compared the semantic similarities between descriptions of the same target based on who produced the description. Different children in the same game produced more similar descriptions than children in different games, although the descriptions were less similar than descriptions from the same child in different rounds (Figure \ref{sim}). We modeled this as a Bayesian mixed effects linear regression predicting similarity.^[`r sim_across_between_form |> form()`] Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`) and were slightly more similar still if they came from the same person with the partnership (`r sim_across_between_1_results |> stats_text(3,3)`). The big differences in descriptions between games compared to within games is a measure of partner sensitivity.

@@

The biggest change between the experiments was increasing the number of repetitions of target stimuli from 3 to 4 (from 12 to 16 trials). The greater number of trials in Experiment 2 made it possible to look for changes over time that could be indicative of convergence to shared descriptions within a game and divergence between games.

### Participants
Experiment 2 was run between March and August of 2024, at the same university preschool as experiment 1. No children participated in both experiments. 30 pairs of children completed all 16 critical trials, and 1 pair of children completed between 8 and 16 critical trials. Our target age range was 4 and 5 year olds, but one older 3-year-old was unintentionally included. Of the 62 children, 30 were girls, and the children had a median age of 56 months, and a range of 45-69 months.

### Materials
The same 4 critical images were used as in experiment 1.In response to some children struggling with the abrupt switch from familiar to non-nameable shapes, we introduced more practice trials for experiment 2. We used a total of 4 practice trials to provide a gradient from familiar shapes to less recognizable, blockier shapes (Figure \ref{game}D).

@@

### Data processing
Data were processed in the same way as experiment 1. After excluding trials where children did not give a description or where the experimenter echoed a child's description, we had `r clean_2 |> filter(!is.na(repNum)) |> nrow()` trials total.

@@

@@

@@

@@

@@

# Joint analysis
